---
title: "DATA 222 - MODEL"
author: "Ritika Lama"
format: pdf
editor: visual
---

```{r dataset}
library(car)
library(randomForest)
library(tidyverse)
apartments <- read_csv("RentWIP_cleaned.csv")
```

## 1. Test-train split

```{r test-train}
set.seed(123)
train <- apartments %>% sample_frac(.70)
test <- anti_join(apartments, train)
```

## 2. Linear Regression Model with STATE

```{r lrm}
lin_model<- lm(monthly_price ~ state, data = train)
summary(lin_model)
```

This linear regression model estimates monthly rent based on state. The intercept of 981.40 represents the average monthly rent for the reference state. Each state coefficient shows how much higher or lower that state’s rent is compared to the reference state. For example, California is about 1815.07 dollars higher, Washington is about 792.99 dollars higher, and New York is about 698.34 dollars higher than the reference state. The model is statistically significant overall, and the R-squared value of 0.2314 means that about 23.14 percent of the variation in monthly rent is explained by state alone.

## 3. Linear Regression Model with square_feet, bedrooms, bathrooms

```{r}
lin_model2<- lm(monthly_price ~ square_feet + bedrooms + bathrooms, data = train)
summary(lin_model2)
```

This multiple linear regression model uses square footage, bedrooms, and bathrooms to predict monthly rent. One additional square foot increases rent by about 0.27 dollars, and one additional bathroom increases rent by about 482.15 dollars. Bedrooms have a small negative effect (-15.53) and are not statistically significant, meaning they do not strongly affect rent after controlling for size and bathrooms. The R-squared value is 0.1572, meaning the model explains about 15.72 percent of the variation in rent.

## 4. Testing with the linear model

```{r linear-test}
linear_test <- predict(lin_model2, newdata = test)
actual <- test$monthly_price
pred <- linear_test

rmse <- sqrt(mean((actual - pred)^2, na.rm = TRUE))
mae <- mean(abs(actual - pred), na.rm = TRUE)

rmse
mae
```

The model performance shows RMSE = 953.77 and MAE = 536.54. This means predictions are off by about 537 dollars on average, and larger errors increase RMSE to about 954 dollars. This indicates the linear model is not very accurate.

## 5. Random Forest Model

```{r random-forest}
train_clean <- na.omit(train)
rf_model <- randomForest(monthly_price ~ ., data = train_clean, ntree = 500)
rf_model
```

Since the Random Forest model included all the predictors from train_clean, it successfully predicted the variable "monthly_price". The type of the response variable is numeric, thus, this model is called a regression model, while not classification. The number of trees grown is 500; this provides more stability and reliability. At each split of the tree, only two randomly chosen predictors were considered. Squared residuals' mean (811328.5) indicates the average squared error of prediction, and smaller values correspond to higher accuracy. In addition, 31.58% of variance in the monthly rent was accounted for by this model, indicating that some information about this variable can be found using predictors, while another share of variance remains dependent on additional factors.

## 6. Testing with Random Forest

```{r rf-test}
test_clean <- na.omit(test)
rf_test <- predict(rf_model, newdata = test_clean)
actual_rf <- test_clean$monthly_price
pred_rf <- rf_test

rmse_rf <- sqrt(mean((actual_rf - pred_rf)^2))
mae_rf <- mean(abs(actual_rf - pred_rf))

rmse_rf
mae_rf
```

The output of this model gave the RMSE value as 700.48, meaning that the prediction made for monthly rent differs by \$700 on average. An MAE value of 399.57 indicates an average error of about \$400 per month. It can be inferred that this model provides reasonable predictions with typical prediction errors ranging from hundreds of dollars.

## 7. CART model

```{r}
library(rpart)
library(rpart.plot)

cart_model <- rpart(
  monthly_price ~ bathrooms+bedrooms+pets_allowed+square_feet,
  data = train_clean
)

cart_model

cart_pred <- predict(cart_model, newdata = test_clean)

cart_rmse <- sqrt(mean((test_clean$monthly_price - cart_pred)^2))
cart_mae <- mean(abs(test_clean$monthly_price - cart_pred))

cart_rmse
cart_mae
```

According to the CART model, the RMSE value is equal to 837.76, meaning that there is an average difference of around \$838 between the forecasted monthly rents and the real values considering that large discrepancies carry more significance. An MAE value of 493.86 indicates that the average error is around \$494 per month. These relatively high error values indicate that the CART model is less precise than the Random Forest model; however, it can still be used for its interpretable nature and identification of significant variables.

```{r}
# Tree Plot
rpart.plot(cart_model)
```

```{r}
# Actual vs Predicted Plot
ggplot(data.frame(
  Actual = test_clean$monthly_price,
  Predicted = cart_pred
), aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.5) +
  geom_abline() +
  labs(
    title = "CART: Actual vs Predicted Monthly Rent"
  ) 
```

The Random Forest model performed best because it had the lowest RMSE and MAE, meaning it gave the most accurate predictions. The CART model was less accurate but easier to interpret. The linear regression models helped explain relationships between variables but were less accurate for prediction. Overall, Random Forest is best for prediction, while CART and linear regression are better for understanding the data.
